Naturalness in U(1) extensions of the MSSM

### 2014-11-11 paper/e6ssm-tuning-paper.tex
updated references, introduction mostly complete.

\title{Naturalness in U(1) exentions of the MSSM}

@@

\begin{abstract}
% Copied SUSY2014 abstract as placeholder
With the discovery of a 125 GeV Standard Model (SM) like Higgs boson
and rising lower bounds on the masses of superpartners, minimal models
of low energy supersymmetry (SUSY) appear to be increasingly fine
tuned. In the Minimal Supersymmetric Standard Model (MSSM) large stop
masses, needed to obtain the observed Higgs mass, give a significant
contribution to fine tuning. $U(1)$ extensions of the MSSM that solve
the $\mu$-problem have new F- and D-term contributions to the Higgs mass,
allowing a 125 GeV Higgs with lighter stops and so alleviating the
associated fine tuning. However, it was recently demonstrated, within
a constrained version of the Exceptional Supersymmetric Standard Model (E$_6$SSM)
defined at the Grand Unification (GUT) scale, that the presence of a massive $Z'$ boson introduces a 
substantial new source of tuning in these theories at tree level. Here we
consider the impact of this tuning when we remove the fine tuning that depends
on assumptions about the SUSY breaking scale, by setting the SUSY
breaking parameters at low energies. We find that the existing
experimental bounds on the $Z'$ mass impose an effective lower bound on
the fine tuning, which does not depend on assumptions about SUSY breaking. 
We compare our results against the tuning in
the phenomenological MSSM (pMSSM), where the SUSY breaking terms are also
defined at low energies.

@@

\section{\label{sec:intro}Introduction}
The discovery of an approximately 125 GeV Higgs
\cite{Aad:2012tfa,Chatrchyan:2012ufa} at the large hadron collider
(LHC) has interesting implications for physics beyond the standard
model (SM) and supersymmetry (SUSY).  On the one hand it provides a
light Higgs boson in fitting with expectations from supersymmetry, and
can be fitted in the minimal supersymmetric standard model (MSSM).  On
the other hand it is slightly heavier than the constrained versions of the 
MSSM (cMSSM) can accommodate naturally \cite{Cassel:2011zd,Ghilencea:2012gz}.

The Higgs causes a naturalness problem because at tree level it has an
upper bound of $m_Z$. The dominant higher order corrections to the
Higgs mass come from stops and to obtain a $125$ GeV Higgs they need
be rather heavy. Heavy stops will provide a large contribution to the
soft breaking mass for the up-type Higgs scalar, $m_{H_u}^2$ at low
energies, through the evolution of the renormalisation group equations
(RGEs) from the grand unification (GUT) scale to the electroweak (EW)
scale. This affects the SUSY prediction of the electroweak VEV, v, or
the mass of the $Z$ boson, $m_Z$.  This naturalness problem motivates
both further examination of non-minimal SUSY models that can raise the
Higgs mass without the need for heavy stops and alternative
possibilities for how the soft breaking parameters get generated,
which might set them them at lower energies, reducing the influence
the stops have on the soft breaking mass for the up-type Higgs scalar,
$m_{H_u}^2$.

In addition to that naturalness issue, often referred to as the little
hierarchy problem, the MSSM also suffers fromn the $\mu$-problem. This
is also a naturalness problem since there should be a natural explanation
of how the $\mu$ superpotential parameter can be set to the same scale
as the soft breaking masses.

U(1) extensions of the MSSM provide a very elegant solution to this
$\mu$-problem \cite{} and also raise the Higgs mass with new $F$- and
$D$-terms. Nonetheless, as was recently demonstrated in the context of
the exceptional supersymmetric standard model (E$_6$SSM)
\cite{King:2005jy,King:2005my,Athron:2010zz}, can still suffer from
naturalness problems with the mass of the new $Z^\prime$ associated
with break down of the new U(1) appearing in the electroweak symmetry
breaking conditions at tree level \cite{Athron:2013ipa}.  Nonetheless
the constrained version of E$_6$SSM (cE$_6$SSM) \cite{Athron:2009ue,
  Athron:2009bs} still turned out to be significantly less tuned than
the cMSSM.

However this comparison of fine tuning depends crucially upon the
assumptions of these gravity mediated SUSY breaking motivated
constrained models and in particular the way the universality
constraints are applied at the GUT scale.  As mentioned above, given
the findings at the LHC it is worth considering other possibilities,
which may allow the soft masses to be set at lower energies. As the
scale at which the parameters fufill some breaking inspired
constraints is lowered the stop masses contribute less to the fine
tuning.

At the same time in U(1) extensions, lowering the UV boundary scale
for the RGE evolution also allows even larger $F$-term contriubutions
to the Higgs mass, so long as one only requires $\lambda$, the
coupling between the Singlet Higgs, $S$ and the up- and down-type
Higgs bosons, $H_u$ and $H_d$, to remain perbative up to the UV scale
and not the all teh up to the GUT scale.

@@

In this paper we investigate how big this tuning is if we bring this
scale all the way down to 20 TeV, effectively minimising contribution
from the stops. We find that the $Z^\prime$ limit is enough to already
require moderate fine tuning.  We also show this is comparable to the
situation in the MSSM defined at smae scalew if chargino limit is
$700$ GeV, which is the most conservative limit on can obtain and
above the tuning from the stops, in both the MSSM and U(1) extensions.
We also show that the tuning from the stops becomes larger when the
high scale is raied to ?? wheares for the E$_6$SSM the minimum tuning
is lower than the MSSM above this scale.  Therefore iot is clear that
which model is preferred and how large the tuning is depends on our
assumptions about SUSY breaking models.

However we also show that if the $Z^\prime$ limit is the extended to
?? then $U(1)$ extensions of the MSSM will already suffer from a
severe tuning, irrespectiove of the UV boundary condition for th RGE
evolution.  Therefore the $Z^\prime$ constraint is amongst the most
important in terms of tuning and attacking natural supersymmtry
experimentally and the next run of the LHC will be crucial.

U(1) literature review  \newline

E6 literature review \newline
 Exceptional Supersymmetric Standard Model is a very well U(1) extension where....

fine tuning literature review... \newline  It has also been shown that the NMSSM is less fine tuned than in it's minimal model rather than with the inclusion of extra matter \cite{Binjonaid:2014oga}.  \newline

summary of paper

\section{\label{sec:model}The Model}

>>>>

\title{Naturalness in U(1) extensions of the MSSM}

@@

\begin{abstract}
% Copied SUSY2014 abstract as placeholder
With the discovery of a 125 GeV Standard Model (SM) like Higgs boson
and rising lower bounds on the masses of superpartners, minimal models
of low energy supersymmetry (SUSY) appear to be increasingly fine
tuned. In the Minimal Supersymmetric Standard Model (MSSM) large stop
masses, needed to obtain the observed Higgs mass, give a significant
contribution to fine tuning. $U(1)$ extensions of the MSSM that solve
the $\mu$-problem have new F- and D-term contributions to the Higgs mass,
allowing a 125 GeV Higgs with lighter stops and so alleviating the
associated fine tuning. However, it was recently demonstrated, within
a constrained version of the Exceptional Supersymmetric Standard Model (E$_6$SSM)
defined at the Grand Unification (GUT) scale, that the presence of a massive $Z'$ boson introduces a 
substantial new source of tuning in these theories at tree level. Here we
consider the impact of this tuning when we remove the fine tuning that depends
on assumptions about the SUSY breaking scale, by setting the SUSY
breaking parameters at low energies. We find that the existing
experimental bounds on the $Z'$ mass impose an effective lower bound on
the fine tuning, which does not depend on assumptions about SUSY breaking. 
We compare our results against the tuning in the MSSM WITH the SUSY breaking terms also defined at low energies.

@@

\section{\label{sec:intro}Introduction}
The discovery of an approximately 125 GeV Higgs
\cite{Aad:2012tfa,Chatrchyan:2012ufa} at the large hadron collider
(LHC) has interesting implications for physics beyond the standard
model (SM) and supersymmetry (SUSY).  On the one hand it provides a
light Higgs boson in fitting with expectations from supersymmetry, and
can be fitted in the minimal supersymmetric standard model (MSSM).  On
the other hand it is slightly heavier than the constrained version of the 
MSSM (cMSSM) can accommodate naturally \cite{Cassel:2011zd,Ghilencea:2012gz}.

The Higgs causes a naturalness problem because at tree level it has an
upper bound of the mass of the $Z$ boson, $m_Z$. The dominant higher
order corrections to the Higgs mass come from stops and to obtain a
$125$ GeV Higgs they need be rather heavy. Heavy stops will provide a
large contribution to the soft breaking mass for the up-type Higgs
scalar, $m_{H_u}^2$ at low energies, through the evolution of the
renormalisation group equations (RGEs) from the grand unification
(GUT) scale to the electroweak (EW) scale. This affects the SUSY
prediction of the electroweak VEV, v, or $m_Z$.  This naturalness
problem motivates both further examination of non-minimal SUSY models
that can raise the Higgs mass without the need for heavy stops and
alternative possibilities for how the soft breaking parameters get
generated, which might set them them at lower energies, reducing the
influence the stops have on the soft breaking mass for the up-type
Higgs scalar, $m_{H_u}^2$.

In addition to that naturalness issue, often referred to as the little
hierarchy problem, the MSSM also suffers from the $\mu$-problem. This
is also a naturalness problem since there should be a natural explanation
of how the $\mu$ superpotential parameter can be set to the same scale
as the soft breaking masses.

U(1) extensions of the MSSM provide a very elegant solution to this
$\mu$-problem \cite{Fayet:1977yc, Kim:1983dt, Suematsu:1994qm, Cvetic:1995rj,  Cvetic:1996mf, Jain:1995cb, Nir:1995bu, Cvetic:1997ky} and also
raise the Higgs mass with new $F$- and $D$-terms. Nonetheless, as was
recently demonstrated in the context of the exceptional supersymmetric
standard model (E$_6$SSM)
\cite{King:2005jy,King:2005my,Athron:2010zz}, such models can still suffer from
naturalness problems with the mass of the new $Z^\prime$ associated
with break down of the new U(1) appearing in the electroweak symmetry
breaking conditions at tree level \cite{Athron:2013ipa}.  Nonetheless
the constrained version of E$_6$SSM (cE$_6$SSM) \cite{Athron:2009ue,
  Athron:2009bs} still turned out to be significantly less tuned than
the cMSSM.

However this comparison of fine tuning depends crucially upon the
assumptions of these gravity mediated SUSY breaking motivated
constrained models and in particular the way the universality
constraints are applied at the GUT scale.  As mentioned above, given
the findings at the LHC it is worth considering other possibilities,
which may allow the soft masses to be set at lower energies. As the
scale at which the parameters fulfill some breaking inspired
constraints is lowered the stop masses contribute less to the fine
tuning.

At the same time in U(1) extensions, lowering the UV boundary scale
for the RGE evolution also allows even larger $F$-term contributions
to the Higgs mass, so long as one only requires $\lambda$, the
coupling between the Singlet Higgs, $S$ and the up- and down-type
Higgs bosons, $H_u$ and $H_d$, to remain perturbative up to the UV scale
and not the all the way up to the GUT scale.

@@

In this paper we investigate how big this tuning is if we bring this
scale all the way down to 20 TeV, effectively minimising contribution
from the stops. We find that the $Z^\prime$ limit is enough to already
require moderate fine tuning.  We also show this is comparable to the
situation in the MSSM defined at the same scale if chargino limit is
$700$ GeV, which is the most conservative limit on can obtain and
above the tuning from the stops, in both the MSSM and U(1) extensions.
We also show that the tuning from the stops becomes larger when the
high scale is raised to ?? whereas for the E$_6$SSM the minimum tuning
is lower than the MSSM above this scale.  Therefore it is clear that
which model is preferred and how large the tuning is depends on our
assumptions about SUSY breaking models.

However we also show that if the $Z^\prime$ limit is the extended to
?? then $U(1)$ extensions of the MSSM will already suffer from a
severe tuning, irrespective of the UV boundary condition for the RGE
evolution.  Therefore the $Z^\prime$ constraint is amongst the most
important in terms of tuning and attacking natural supersymmetry
experimentally and the next run of the LHC will be crucial.

Here we will refer to U(1) extensions of the MSSM, which solve the
$\mu$-problem as the USSM
\cite{Cvetic:1995rj, Jain:1995cb, Nir:1995bu, Cvetic:1996mf, Cvetic:1997ky}, while we will
sometimes specify that the models are inspired by $E_6$ by referring
to them as such. Finally we will also discuss specific models with
complete $E_6$ matter multiplets at low energies and an extra $U(1)$
under which the right handed neutrino remains interaction-less as
E$_6$SSM-variants\cite{Howl:2007zi, Braam:2009fi, Braam:2010sy, Hall:2011zq, Nevzorov:2012hs, Athron:2014pua}, with the E$_6$SSM reserved for the original
formulation of this model \cite{King:2005jy,King:2005my,Athron:2010zz}
original formulation of the model.

$U(1)$ and $E_6$ inspired extensions of the MSSM have been studied
very widely in the literature \cite{Gunion:1989we, Gunion:1992hs,
  Binetruy:1985xm, Ellis:1986yg, Ibanez:1986si, Gunion:1986ky,
  Haber:1986gz, Baer:1987eb, Gunion:1987jd, Grifols:1986vr,
  Ellis:1986ip, Morris:1987fm, Drees:1987tp, Ma:1995xk,
  Suematsu:1997tv, Suematsu:1997qt, Suematsu:1997au, Keith:1996fv,
  Keith:1997zb, Gherghetta:1996yr, Demir:1998dk, Langacker:1998tc,
  Hambye:2000bn, Ma:2000jf} (or for reviews see
\cite{Hewett:1988xc,Langacker:2008yv}).  There has also been a lot of
work recently including investigations of the neutralino sector
\cite{Hesselbach:2001ri, Barger:2005hb, Choi:2006fz,
  Barger:2007nv};the relic abundance of dark matter
\cite{Kalinowski:2008iq}; GUT scale family symmetries which can
explain the hierarchy of masses in the fermion sector and their
associated mixings \cite{Stech:2008wd}; neurtrino physics
\cite{Kang:2004ix}; explanations of the matter-anti-matter asymmetry
of the universe though EW baryogensis or leptogensis
\cite{Hambye:2000bn,Ma:2000jf,Kang:2004pp}; dipole moments
\cite{GutierrezRodriguez:2006hb}; anomaly mediated SUSY breaking with
D-term contributions \cite{Asano:2008ju} and the (extended) Higgs
sectors \cite{Daikoku:2000ep,Ham:2008xf}.

In this paper we will often focus on a very well motivated $E_6$
inspired $U(1)$ extension, the Exceptional Supersymmetric Standard
Model (E$_6$SSM)\cite{King:2005jy,King:2005my,Athron:2010zz}.  In this
model since the right handed neutrino remains interaction-less they may
explain the tiny observed masses of neutrinos via the see-saw
mechanism and the baryon asymmetry in the universe via leptogenesis
\cite{Hambye:2000bn,King:2008qb, King:2008gw}.

The gauge coupling running in the E$_6$SSM at the two-loop level leads
to unification more precisely than in the MSSM \cite{King:2007uj} or
in slightly modified scenarios two-step unification can take place
\cite{Howl:2007hq,Howl:2007zi}. If the exotic particles are light in
these models this can open up non-standard decays of the SM--like
Higgs boson \cite{Hall:2010ix,Nevzorov:2013tta,Athron:2014pua}. The
correct relic abundance could be obtained entirely through an almost
decoupled ``inert'' neutralino sector \cite{Hall:2009aj},.  However
this is no longer phenomenologically viable due to limits from
direct detection of dark matter
\cite{2011PhRvL.107m1302A,2012PhRvL.109r1301A, Akerib:2013tjd} and due
to a significant suppression of the decay of the lightest Higgs boson
into SM states, due to a new channel into inert singlinos opening up.

There are still several remaining options. One may specialize to
scenarios known as the EZSSM \cite{Hall:2011zq} where the inert
singlinos, which cause these problems are entirely decoupled and the
relic abundance is fitted with a bino-like candidate with a novel
mechanism involving back-scattering into a Heavier inert
Higgsino. Another well motivated scenario admits two possible dark
matter candidates \cite{Nevzorov:2012hs}, where one will be an inert
singlino and the other will have a similar composition to MSSM
neutralinos. The simplest phenomenologically viable solution in that
case is to make the singlinos extremely light hot dark matter
candidate and explains almost all the relic abundance density with

The impact of gauge kinetic
mixing in the case where both of the extra U(1)'s appearing from the
breakdown of E$_6$ are present at low energy was studied in
\cite{Rizzo:2012rf}.  The E$_6$SSM was also included in a studies looking
at how first or second generation sfermion masses can be used to
constrain the GUT scale parameters \cite{Miller:2012vn} and the
renormalization of VEVs \cite{Sperling:2013eva, Sperling:2013xqa}.
The particle spectrum and  collider signatures of constrained version of the E$_6$SSM have been studied in a series of papers,
\cite{Athron:2009ue,Athron:2009bs,Athron:2011wu,Athron:2012sq}. The threshold
corrections to the $\overline{DR}$ gauge and Yukawa couplings in the
E$_6$SSM were calculated and the numerical impact in the constrained
version examined \cite{Athron:2012pw}.

As mentioned earlier the fine tuning of the cE$_6$SSM was recently studied
\cite{Athron:2013ipa} and there it was revealed that the associated
$Z^\prime$ boson leads to a new source of fine tuning since it's mass
appears in the EWSB conditions.

In this paper we will examine this source of fine tuning in more detail by considering low energy constructions where the usual fine tuning problem from the Higgs is minimised.   We will also consider  alternate charges for the extra $U(1)$ symmetry to relax the focus on the E$_6$SSM and demonstrate that this is quite a generic result.

To quantify the fine tuning we will employ the traditional
Barbieri-Giudice measure \cite{Ellis:1986yg,Barbieri:1987fn}.  This has
been used extensively within the literature
e.g. \cite{deCarlos:1993yy, deCarlos:1993ca, Chankowski:1997zh, Agashe:1997kn, Wright:1998mk, Kane:1998im, BasteroGil:1999gu, Feng:1999zg, Allanach:2000ii, Dermisek:2005ar, Barbieri:2005kf, Allanach:2006jc, Gripaios:2006nn, Dermisek:2006py, Barbieri:2006dq, Kobayashi:2006fh, Perelstein:2012qg, Antusch:2012gv,Cheng:2012pe, CahillRowley:2012rv, Ross:2012nr, Basak:2012bd, Kang:2012sy, Athron:2013ipa,Miller:2013jra, Binjonaid:2014oga, Miller:2014jza}.

A number of alternative measures have also been applied in the
literature \cite{Anderson:1994dz, Anderson:1994tr,
  Anderson:1995cp,Anderson:1996ew,Ciafaloni:1996zh,Chan:1997bi,
  Barbieri:1998uv,
  Giusti:1998gz,Casas:2003jx,Casas:2004uu,Casas:2004gh,Casas:2006bd,Kitano:2005wc,
  Athron:2007ry, Athron:2007qr, Baer:2012up} with varying motivations.
A very different approach is to work within a Bayesian analysis.
There the concept of naturalness is automatically incorporated since
in models where one must fine tune parameters to fit measured values
of the observables, the region with high likelihood will occupy a tiny
prior volume \cite{Allanach:2007qk, Cabrera:2008tj, Ghilencea:2012gz,
  Ghilencea:2012qk, Fichet:2012sn, Kim:2013uxa} suppressing the
posterior.  Indeed in the MSSM and NMSSM if one transforms GUT scale
parameters to the VEVs, the inverse of the Jacobian for this
transformation looks quite like the derivatives that appear in the
traditional fine tuning measure \cite{Allanach:2007qk, Cabrera:2008tj,
  Kim:2013uxa}.  If one thinks more generally then a model without
fine tuning is one where the parametrisation is such that all the
parameters are observables \cite{Fichet:2012sn, Kim:2013uxa}.  This
provides a quite general definition of fine tuning as $1/|J|$ where
$|J|$ is the determinant of the Jacobian for the co-ordinate
transformation between the parameters and the
observables. Interestingly this means the tuning is the ratio of the
infinitesimal observable space space volume element to the
infinitesimal parameter space element and coincides with the measure
proposed in \cite{Athron:2007ry} when the interval of variation is
taken to zero.

While this approach has many merits here will employ the traditional
measure of fine tuning because it is both simple to apply and easy top
compare with previous results due to it's widespread use.  Fortunately
the derivatives which appear in these tuning are also similar to the
Bayesian motivated measure so there should not be too large a discrepancy between the two approaches.

The structure of this paper is as follows.  In section \ref{sec:model} we review the models we consider.  In section \ref{sec:ewsb} we specify the EWSB conditions of the models, with particular focus on how the $Z^\prime$ mass influences the prediction of $M_Z$.  Then in section \ref{sec:tuning} we introduce our fine tuning measure and our approach to evaluating it to obtain the individual sensitivities.  The numerical procedure we use to obtain our results is given in \ref{subsec:numericalprocedure} and then the results are specified in section \ref{sec:results}.

\section{\label{sec:model}U(1) extensions and the E$_6$SSM}
